### 2014-06-03 writing/S'14 paper/intro.tex
Revised recursion section

Supervised recursive neural network models (RNNs) for sentence meaning
have been successful in a wide array of sophisticated language tasks,
including sentiment analysis \cite{socher2011semi,socher2013acl1},
image description \cite{sochergrounded}, and paraphase detection
\cite{Socher-etal:2011:Paraphrase}. These results are encouraging
about the ability of these models to learn compositional semantic
grammars, but it remains an open question whether they can achieve the
same results as grammars based in logical forms when it comes to core
semantic concepts like quantification, entailment and contradiction.
To date, experimental investigations of these concepts using
distributed representations have been largely confined to short
phrases
\cite{Mitchell:Lapata:2010,Grefenstette-etal:2011,baroni2012entailment,kalchbrenner2014convolutional}.
For robust natural language understanding, we must be able to model
these phenomena in full generality on complex linguistic structures.

>>>>

Supervised recursive neural network models (RNNs) for sentence meaning
have been successful in a wide array of sophisticated language tasks,
including sentiment analysis \cite{socher2011semi,socher2013acl1},
image description \cite{sochergrounded}, and paraphase detection
\cite{Socher-etal:2011:Paraphrase}. These results are encouraging
about the ability of these models to learn compositional semantic
grammars, but it remains an open question whether they can achieve the
same results as grammars based in logical forms
\cite{ZetCol:2005,LiangJordan:2013} when it comes to core semantic
concepts like quantification, entailment and contradiction. To date,
experimental investigations of these concepts using distributed
representations have been largely confined to short phrases
\cite{Mitchell:Lapata:2010,Grefenstette-etal:2011,baroni2012entailment,kalchbrenner2014convolutional}.
For robust natural language understanding, we must be able to model
these phenomena in full generality on complex linguistic structures.

### 2014-06-03 writing/S'14 paper/recursion.tex
Revisions focused on the join table section
Revised recursion section

\begin{figure}[t]
\begin{center}
\begin{tabular}{lll}
$a\equiv a$		&~~~&	$(c~(and~(not~d)))~\#~f$\\
$b~\#~c$			&~~~&	$(not~(c~(or~b)))~\sqsubset~(not~c)$\\
$d\natneg(not~d)$	&~~~&	$f~\#~((c~(or~(not~d)))~(and~a))$\\
$(c~(and~d))\sqsubset d$&~~~&$d\sqsupset((d~(or~d))~(and~(not~b)))$\\
\end{tabular}
\end{center}

\caption{Some sample randomly generated pairs of propositional logic statements.  \label{prop-figure}} 
\end{figure}

% TODO: Cite Chomsky/Hauser/Fitch?

Recursive structure is a prominent  feature of natural language. Consider, for example, \ii{Alice said hello}, \ii{Bob said that Alice said hello}, and \ii{Carl thinks that Bob said that Alice said hello}. Overt recursion of this kind is easy to find, and theoretical accounts of natural language syntax and semantics rely heavily on recursive structures. In order for a model to be able to accurately learn natural language meanings, then, we expect that it would need to be able to learn to represent the meanings of function words in a such a way that they are able to behave correctly when taking their own outputs as input. In evaluating the model, we take advantage of the fact that recursive structures of this kind define potentially infinite languages by testing the model on strings that are longer and more complex than any seen in testing.

We again test this phenomenon within the framework of MacCartney and Manning-style entailment reasoning, but we replace the unanalyzed symbols from the previous experiment with expressions that involve recursive structure. To define these expressions, we turn to propositional logic, a relatively simple logic in which each variable represents either \ii{true} or \ii{false}. We generate data of the form seen in Figure \ref{prop-figure}: strings of arbitrary length consisting of six elementary proposition variables and the operators \ii{and}, \ii{or}, and \ii{not}, arranged in pairs with the logical relations between them specified.

% NOTE: There was some confusion at CSLI and the 224U lecture about whether this was theorem proving or some kind of model checking. Hopefully this is explicit enough.
Socher et al. \cite{socher2012semantic} have previously demonstrated the learning of a logic in a matrix-vector RNN model somewhat similar to our own, but the logic discussed here is substantially stronger, and a much better approximation of the kind of structure that is needed for natural language. The logic learned in that experiment is boolean, wherein the atomic symbols are simply the values 0 and 1, rather than variables over those values. While learning the operators of that logic is not trivial, the ouptuts of each operator can be represented accurately by a single bit. The statements of propositional logic learned here describe conditions on the truth values of propositions where those truth values are not known. As opposed to the two-way contrasts seen in \cite{socher2012semantic}, this logic distinguishes between 64 (2^6) possible assignments of truth values, and expressions of this logic define arbitrary conditions on these possible assignments, for a total of 2^{64} ($\approx 10^{20}$) possible statements that the recursive model needs to be able to distinguish. To frame this distinction in another way, the relational statements in our data our theorems about the relations between statements in the logic tested in \cite{socher2012semantic}.

We randomly generate pairs of parentheses-bracketed statements of the logic and then randomly divide the results into training and test data sets. To compute the relation between each pair of statements, we exhaustively enumerate the sets of assignments of truth values to proposition variables that would satisfy each of the statements and then convert the set-theoretic relation between those assignments into one of the seven relations. If we do not implement any constraint that the two statements being compared are similar in any way, the generated data consists in large part of statements in which the two refer to largely separate subsets of the six variables, and to which we will nearly always assign the \# relation. In an effort to balance the distribution of relation labels without departing from the basic task of modeling propositional logic, we disallow individual pairs of statements from referring to more than four of the six proposition variables.

We deduplicate the data and discard pairs in which either statement is a tautology or contradiction (a statement that is true of either all or no possible assignments), for which none of the seven relation labels can accurately apply. We then divide the generated pairs into size bins based on the number of logical operators (\ii{and}, \ii{or}, or \ii{not}) in the larger of the two pairs being compared, and discard examples of size greater than twelve by this measure. Finally, we randomly sample 15\% of each bin for a held out test set.

We trained both the RNN and RNTN models on the data of size four or less (65k pairs), and tested it on examples of up to size 12 (44k pairs). We initialized the model parameters randomly, including the vector representations of the six variables.

The results are shown in Figure \ref{prop-results}. In tuning, we found that the RNN model was approximately optimal with 45 dimensional vector representations, and the RNTN model was approximately optimal with 25 dimensions. We fixed the size of the feature vector for the classifier at 75 dimensions. We found that the RNTN model was able to perform almost perfectly on unseen small test examples, with accuracy above 99\% below size four. After depth four, performance gradually falls with increasing size. The RNN model did not perform well, reaching only 88.2\% accuracy on the smallest test examples, and declining from there to near-baseline performance at size 12.

\begin{figure}[t]
\begin{center}
\includegraphics[width=5.5in]{recursion\string_results.eps}
\end{center}

\caption{Model performance by expression size.  \label{prop-results}} 
\end{figure}

>>>>

% TODO: Cite Chomsky/Hauser/Fitch? -- no, not appropriate here --- CP.

Recursive structure is an prominent feature of natural languages and
an important part of their expressivity, as it allows people to use
and interpret complex structures they have never encountered before.
Theoretical accounts of natural language syntax and semantics
therefore rely heavily on recursive structures, and we expect that
accurate computational models will have to come to grips with it as
well. The present section pursues this question for our two classes of
RNN. In evaluating the model, we take advantage of the fact that
recursive structures of this kind define potentially infinite
languages, by testing the model on strings that are longer and more
complex than any seen in testing.

% Consider, for example, \ii{Alice said hello}, \ii{Bob said that Alice
%   said hello}, and \ii{Carl thinks that Bob said that Alice said
%   hello}. Overt recursion of this kind is easy to find, and
% theoretical accounts of natural language syntax and semantics rely
% heavily on recursive structures. In order for a model to be able to
% accurately learn natural language meanings, then, we expect that it
% would need to be able to learn to represent the meanings of function
% words in a such a way that they are able to behave correctly when
% taking their own outputs as input. In evaluating the model, we take
% advantage of the fact that recursive structures of this kind define
% potentially infinite languages by testing the model on strings that
% are longer and more complex than any seen in testing.

As in Section~\ref{sec:join}, we test this phenomenon within the
framework of natural logic, but we now replace the unanalyzed symbols
from that experiment with expressions that involve recursive
structure. The result is a truth-functionally complete classical
propositional logic: each atomic variable denotes either $\True$ or
$\False$, and the only operators are truth-functional ones.
Table~\ref{tab:pl} defines this logic, and Table~\ref{tab:plexs} gives
some examples of natural logic relational statements that we can make
in these terms. To compute these relations between statements, we
exhaustively enumerate the sets of assignments of truth values to
proposition variables that would satisfy each of the statements and
then convert the set-theoretic relation between those assignments into
one of the seven relations in Table~\ref{b-table}. This is equivalent
to proving theorems about these statements model-theoretically.

\newcommand{\plneg}{\mathop{\textit{not}}}
\newcommand{\pland}{\mathbin{\textit{and}}}
\newcommand{\plor}{\mathbin{\textit{or}}}

\begin{table}[htp]
  \centering
  \begin{subtable}[t]{0.45\textwidth}
    \centering
    \begin{tabular}[t]{l l}
      \toprule
      Formula     & Interpretation \\
      \midrule
      $a$, $b$, $c$, $d$, $e$, $f$ & $\sem{x} \in \{\True, \False\}$ \\
      $\plneg \varphi$ & $\True$ iff $\sem{\varphi} = \False$ \\
      $(\varphi \pland \psi)$ & $\True$ iff $\False \notin \{\sem{\varphi}, \sem{\psi}\}$ \\
      $(\varphi \plor \psi)$  & $\True$ iff $\True \in \{\sem{\varphi}, \sem{\psi}\}$ \\
      \bottomrule
    \end{tabular}    
    \caption{Well-formed formulae. $\varphi$ and $\psi$
      range over all well-formed formulae, and $\sem{\cdot}$ is
      interpretation function mapping formulae into $\{\True,
      \False\}$.}\label{tab:pl}
  \end{subtable}
  \quad
  \begin{subtable}[t]{0.45\textwidth}
    \centering
    \begin{tabular}[t]{r c l}
      \toprule
      $\plneg a$        & $\natneg$ & $a$ \\
      $\plneg \plneg a$ & $\nateq$  & $a$ \\
      $a$               & $\natfor$ & $(a \plor b)$ \\
      $a$               & $\natrev$ & $(a \pland b)$ \\
      $(a \natfor b)$   & $\nateq$  & $(b \natrev a)$ \\	
      $\plneg(\plneg a \pland \plneg b)$ & $\nateq$ & $(a \plor b)$ \\ 
      \bottomrule
    \end{tabular}
    \caption{Examples of statements about relations between
      well-formed formulae, defined in terms of sets of satisfying
      interpretation functions $\sem{\cdot}$.}\label{tab:plexs}
  \end{subtable}
  \caption{Propositional logic with natural logic relations.}  
  \label{prop-figure}
\end{table}

% NOTE: There was some confusion at CSLI and the 224U lecture about
% whether this was theorem proving or some kind of model checking.
% Hopefully this is explicit enough.
Socher et al.~\cite{socher2012semantic} demonstrate the learning of a
logic in a matrix-vector RNN model somewhat similar to our own, but
the logic discussed here is substantially stronger, and a much better
approximation of the kind of structure that is needed for natural
language. The logic learned in that experiment is boolean, wherein the
atomic symbols are simply the values $0$ and $1$, rather than
variables over those values. While learning the operators of that
logic is not trivial, the outputs of each operator can be represented
accurately by a single bit. The statements of propositional logic
learned here describe conditions on the truth values of propositions
where those truth values are not known. As opposed to the two-way
contrasts seen in \cite{socher2012semantic}, this logic distinguishes
between $2^{6} = 64$ possible assignments of truth values, and
expressions of this logic define arbitrary conditions on these
possible assignments, for a total of $2^{64}$ %($\approx 10^{20}$)
possible statements that the recursive model needs to be able to
distinguish.

For our experiments, we randomly generate large sets of syntactically
unique formulae and compute the natural logic relation between them.
We discard pairs in which either statement is a tautology or
contradiction (a statement that is true of either all or no possible
assignments), for which none of the seven relation labels in
Table~\ref{b-table} can hold. The resulting set of formula pairs is
then partitioned into bins based on the number of logical operators in
the longer of the two formulae. We then randomly sample 15\% of each
bin for a held out test set.

If we do not implement any constraint that the two statements
being compared are similar in any way, the generated data consists in
large part of statements in which the two formulae refer to largely
separate subsets of the six variables, and to which we will nearly
always assign the $\natind$ relation. In an effort to balance the
distribution of relation labels without departing from the basic task
of modeling propositional logic, we disallow individual pairs of
statements from referring to more than four of the six proposition
variables.

% We deduplicate the data and discard pairs in which either statement is
% a tautology or contradiction (a statement that is true of either all
% or no possible assignments), for which none of the seven relation
% labels can accurately apply. We then divide the generated pairs into
% size bins based on the number of logical operators (\ii{and}, \ii{or},
% or \ii{not}) in the larger of the two pairs being compared, and
% discard examples of size greater than twelve by this measure. Finally,
% we randomly sample 15\% of each bin for a held out test set.

We trained both the RNN and RNTN models on the data of size four or
less (65k pairs), and tested it on examples of up to size 12 (44k
pairs). We initialized the model parameters randomly, including the
vector representations of the six variables. The results are shown in
Figure~\ref{prop-results}. In tuning, we found that the RNN model was
approximately optimal with 45-dimensional vector representations, and
the RNTN model was approximately optimal with 25 dimensions. We fixed
the size of the feature vector for the classifier at 75 dimensions. We
found that the RNTN model was able to perform almost perfectly on
unseen small test examples, with accuracy above 99\% below size four.
After depth four, performance gradually falls with increasing size.
The RNN model did not perform well, reaching only 88.2\% accuracy on
the smallest test examples, and declining from there to near-baseline
performance at size 12.

\begin{figure}[htp]
  \centering
  \includegraphics[width=5.5in]{recursion\string_results.eps}
  \caption{Model performance by expression size.}  
  \label{prop-results}
\end{figure}
